Symbolic Dynamic Programming for Continuous State and Observation POMDPs

### 2012-11-01 cpomdp/camdp-xadd/doc/NIPS12/draft6.tex
Intermediate checkin

% TODO: 
%
% (1) GenRelObs and propagate fixes to text
%- i's where they shouldn't be, 
%- integrating over discrete vars, 
%- P(z_k|...) needs to condition on belief state, 
%- line 13 generates does not get rid of xd_s'... did you mean to integrate out alpha(xd_o,xd_s) instead?
% ... \delta_j(x_{d_o}) should integrate out to 1 right?
%     then use this?
%
% (2) Notation for observation set
%
% (3) Improve intuitions
%
% (4) Revise conclusion
%
% (5) More formal claims, where this works, where PBVI exact

@@

\def\argmax{\operatornamewithlimits{arg\max}}

@@

\author
{
Zahra Zamani \\
NICTA & ANU\\
Canberra, Australia \\
\texttt{zahra.zamani@anu.edu.au} \\
\And
Scott Sanner \\
NICTA & ANU\\
Canberra, Australia \\
\texttt{ssanner@nicta.com.au} \\
\And
Pascal Poupart \\
University of Waterloo \\
Waterloo, Canada \\
\texttt{ppoupart@uwaterloo.ca} \\
\And
Kristian Kersting \\
Fraunhofer IAIS, Sankt Augustin, Germany \\
IGG, University of Bonn, Germany \\
\texttt{kristian.kersting@iais.fraunhofer.de} \\
}

@@

\newcommand{\open}{\mathit{open}}
\newcommand{\close}{\mathit{close}}
\newcommand{\high}{\mathit{high}}
\newcommand{\low}{\mathit{low}}
\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}

%\nipsfinalcopy % Uncomment for camera-ready version

@@

\begin{abstract}
Partially-observable Markov decision processes (POMDPs) provide a
powerful model for real-world sequential decision-making problems.  In
recent years, point-based value iteration methods have proven to be
extremely effective techniques for finding (approximately) optimal
dynamic programming solutions to POMDPs when an initial set of belief
states is known.  However, no point-based work has provided exact
point-based backups for both continuous state and observation spaces,
which we tackle in this paper.  Our key insight is that while there
may be an infinite number of possible observations, there are only a
finite number of observation partitionings that are relevant for
optimal decision-making when a finite, fixed set of reachable belief
states is known.  To this end, we make two important contributions:
(1) we show how previous exact symbolic dynamic programming solutions
for continuous state MDPs can be generalized to continuous state
POMDPs with discrete observations, and (2) we show how this solution
can be further extended via recently developed symbolic methods to
continuous state and observations to \emph{derive} the minimal
relevant observation partitioning for potentially correlated,
multivariate observation spaces.
%for domais with piecewise
%polynomial transition and observation functions.  
We demonstrate our algorithm on a power plant
control problem requiring continuous state 
\emph{and} observations.
%empirically evaluate on real-world inspired power 
%proof-of-concept results on uni- and multi-variate state and
%observation power plant control.
\end{abstract}

@@

Partially-observable Markov decision processes (POMDPs) are a powerful
modeling formalism for real-world sequential decision-making
problems~\cite{kaebling}.  In recent years, point-based value
iteration methods~\cite{pbvi_jair06,hsvi2,Perseus,gapmin} have proved
extremely successful at scaling (approximately) optimal POMDP
solutions to large state spaces when a set of initial belief states is
known.

More recent work has extended point-based methods to both continuous
state and continuous observation spaces, but no work has tackled both
jointly without sampling.  \cite{Perseus_cont} provides exact
point-based backups for continuous state and discrete observation
problems (with approximate sample-based extensions to continuous
actions and observations), while~\cite{pascal_ijcai05} provides exact
point-based backups for discrete state and continuous observation
problems (where multivariate observations must be conditionally
independent).  While
restricted to discrete states, \cite{pascal_ijcai05} provides an
important insight that we exploit in this work: \emph{only a finite
number of partitionings of the observation space are required to
distinguish between the optimal conditional policy over a finite set
of belief states}.

Our contributions in this work are two-fold.  First, we extend
symbolic dynamic programming for continuous state
MDPs~\cite{sanner_uai11} to the case of continuous state and discrete
observation POMDPs.  This provides an expressive and concrete
instantiation of the framework in~\cite{Perseus_cont} (which only
abstractly required that integrals were computable) in that it ensures
the closed-form of all $\alpha$-functions for \emph{all horizons} is a
symbolic piecewise case statement, even for POMDPs with
\emph{arbitrary} continuous rewards and transitions with discrete
noise (i.e., a finite mixture of deterministic transitions).  
Second, we extend this symbolic dynamic programming algorithm
to the case of continuous observations (while restricting the
transition to piecewise linear functions and the reward functions to piecewise constants) and extend
the work of~\cite{pascal_ijcai05} to \emph{derive} relevant observation
partitions for potentially correlated multivariate continuous
observation spaces by exploiting the piecewise polynomial integration
operation of~\cite{sanner_aaai12} and the multivariate symbolic maximization
technique of~\cite{sanner_uai11}.  
We conclude by demonstrating our algorithm on a power plant
control problem requiring bivariate continuous state and observations.

%\footnote{All code for these
%experiments is freely available online (link suppressed) and
%attached as anonymized supplementary material to this submission.}

@@

We assume familiarity with MDPs and introduce discrete and continuous
partially observable MDPs (DC-POMDPs) as an extension to their DC-MDP
subset~\cite{sanner_uai11}.  A DC-POMDP is a tuple $\langle
\mathcal{S},\mathcal{A},\mathcal{O},\mathcal{T},\mathcal{R},\mathcal{Z},\gamma,h
\rangle$.  States $\mathcal{S}$ are represented by vectors of
$(\vec{d_s},\vec{x_s}) = (
d_{s_1},\ldots,d_{s_n},x_{s_1},\ldots,x_{s_m} )$ where each $d_{s_i}
\in \{ 0,1 \}$ ($1 \leq i \leq n$) is a discrete boolean and each
$x_{s_j} \in \mathbb{R}$ ($1 \leq j \leq m$) is a continuous variable.
We assume a finite, discrete action space $\mathcal{A} = \{ a_1,
\ldots, a_p \}$. Observations
$\mathcal{O}$ are given by the vector $(\vec{d_o},\vec{x_o}) = (
d_{o_1},\ldots,d_{o_p},x_{o_1},\ldots,x_{o_q} )$ where each $d_{o_i}
\in \{ 0,1 \}$ ($1 \leq i \leq p$) is boolean and each $x_{o_j} \in
\mathbb{R}$ ($1 \leq j \leq q$) is continuous.

Three functions are required for modeling DC-POMDPs: (1) $\mathcal{T}: \mathcal{S} \times \mathcal{A} \times \mathcal{S} \rightarrow  [ 0, 1 ]$ a Markovian transition model defined as the probability of the next state given the action and previous state%$P(\vec{d}',\vec{x}'|\cdots,a)$
; (2)  $\mathcal{R}:\mathcal{S}\times\mathcal{A} \rightarrow \mathbb{R}$ a reward function which returns the immediate reward of taking an action in some state; and (3) an observation function defined as $\mathcal{Z} : \mathcal{S} \times \mathcal{A} \times \mathcal{O} \rightarrow [ 0, 1 ]$  which gives the probability of an observation given the outcome of a state after executing an action.  A discount factor $\gamma, \; 0 \leq \gamma \leq 1$ is used to discount rewards $t$ time steps into the future by $\gamma^t$.

We use a dynamic Bayesian network (DBN) to compactly represent the transition model over the factored state variables and we likewise use a Bayesian network to represent the observation model:
{\footnotesize
\begin{align*}
\mathcal{T}: \;\; &
P(\vec{d_s}',\vec{x_s}'|\vec{d_s},\vec{x_s},a) = 
\prod_{i=1}^n P(d_{s_i}'|\vec{d_s},\vec{x_s},a) \prod_{j=1}^m P(x_{s_j}'|\vec{d_s},\vec{d_s}',\vec{x_s},a). \nonumber \\
\mathcal{Z}: \;\; & 
P(\vec{d_o},\vec{x_o}|\vec{d_s},\vec{x_s},a) = 
\prod_{i=1}^n P(d_{o_i}|d_{s_i},a) \prod_{j=1}^m P(x_{o_j}|x_{s_j},a). \nonumber 
\end{align*}}
For simplicity of algorithmic exposition, we do not allow synchronic
arcs in the transition representation, but note that to incorporate
such arcs only requires restrictions on the variable elimination
ordering during dynamic programming.  In these transition and
observation Bayes nets, 
probabilities over 
\emph{discrete} variables are represented using conditional probability
functions (CPF) in the form of
$P(d_i'|\vec{d},\vec{x},a)$. Probabilities for \emph{continuous} variables
$P(x_j'|\vec{d},\vec{d'},\vec{x},a)$ are represented by
\emph{piecewise linear functions} (PLFs) that are first-order Markov
and deterministic, i.e. the probabilities are encoded using the Dirac
$\delta[\cdot]$ function. The reward function $R(\vec{d},\vec{x},a)$
is set to a piecewise constant function. We next provide an intuitive
example to make this DC-POMDP framework more concrete.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 1 - policy tree
\begin{figure}[t!]
%\vspace{-1mm}
\begin{center}
\includegraphics[width=0.4\textwidth]{pics/cond_plan2.pdf}
\hspace{5mm}
\includegraphics[width=0.45\textwidth]{pics/dag2.pdf}
\end{center}
\vspace{-2mm}
\caption{\footnotesize (left) Example conditional plan $\beta^h$ for discrete observations; (right) an example piecewise $\alpha$-vector for continuous state $x$ represented
as a 
%The maximum $\alpha$ vector for a belief state at iteration 4 for the \textsc{\bf Power Plant} problem in Section~\ref{}
decision diagram: 
the \emph{true} branch is solid, the \emph{false}
branch is dashed.}
\label{fig:cond_plan}
\vspace{-1mm}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{figure}[t]
%\begin{center}
%\includegraphics[width=0.4\textwidth]{pics/cond_plan2.pdf}
%\end{center}
%\vspace{-3mm}
%\caption{%\footnotesize 
%The optimal value function for \textsc{\bf Power Plant}
%as a decision diagram: 
%the \emph{true} branch is solid, the \emph{false}
%branch is dashed.} 
%\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\textbf{Example} \textsc{\bf (Power Plant)~\cite{steam2}} \emph{The steam
generation system of a power plant aims to provide hot steam to the
steam turbine which in turn produces electricity. Feed-water is
exposed to heat in the tubes leading to the water tank, where water
evaporation is performed under specific pressure and temperature.
A reward is obtained when electricity is generated through steam
production and the pressure and temperature are within safe ranges.
Mixing water and steam makes the
respective pressure and temperature observations $p_o \in \mathbb{R}$
and $t_o \in \mathbb{R}$ on the underlying state $p_s \in \mathbb{R}$
and $t_s \in \mathbb{R}$ highly uncertain.  Action choices are either
to open or close a pressure valve $a \in \{ \open, \close \}$.}

We initially present two 
simple examples labeled \textsc{\bf 1D-Power Plant} 
using only the temperature state $t_s$.
First, the transition and reward functions common to both problems
are defined as below where simply, the temperate increments or
decrements according to the action:
% For the transition probability, we use the Dirac function to model the deterministic equations. We can also define stochasticity in the transition model using boolean random variables that are sampled stochastically.  The reward function can be any linear function of the state or action. Going above a threshold temperature (e.g. $T=10$) will cause an explosion in the plant when trying to close the valve which results in the negative reward of $-1000$. Staying below this temperature is safe and will produce electricity an gain the reward of $100$. The reward of an open valve is $-1$.
{\footnotesize
\vspace{-1mm}
\begin{align}
\label{eq:trans}
P(t_s'|\vec{t_s},a)= \delta\left[ t_s' - 
\begin{cases}
 (a=\open) &: t_s - 5 \\ 
(a \neq \open) &: t_s + 7 \\
\end{cases}
\right]
\hspace{2mm}
R(t_s,a) = 
\begin{cases}
 (a=\open) &: -1 \\
(a \neq \open)\wedge (t_s>T) &: -1000 \\
(a \neq \open)\wedge \neg(t_s>T) &: 100 \\
\end{cases}
\end{align}
\vspace{-4mm}
}

%contribution? 
Next we introduce the \textsc{\bf Discrete Obs. 1D-Power Plant} variant using 
a discrete binary observation space with possible
observations $o \in \mathcal{O} = \{\high,\low\}$:
{\footnotesize
\vspace{-1mm}
\begin{align}
P(o=\high|t_s',\open) = 
\begin{cases}
  t_s' <= 15 &: 0.9 \\
 \neg (t_s' <= 15)&: 0.1 \\
\end{cases}
\hspace{5mm}
P(o=\low|t_s',\open) = 
\begin{cases}
 t_s' <= 15 &: 0.1 \\
 \neg (t_s' <= 15) &: 0.9 \\
\end{cases}
\end{align}
\vspace{-3mm}
}
%\\
%P(t_{o_1}|t_s',\close) = 
%\begin{cases}
% \delta\left[ t_s <= 15 \right] &: 0.1 \\
% \delta\left[ \neg (t_s <= 15)\right] &: 0.8 \\
%\end{cases}\nonumber
%,
%P(t_{o_2}|t_s',\close) = 
%\begin{cases}
% \delta\left[ t_s <= 15 \right] &: 0.9 \\
% \delta\left[ \neg (t_s <= 15)\right] &: 0.2 \\
%\end{cases} \nonumber
%\end{align}

Finally we introduce the \textsc{\bf Cont. Obs. 1D-Power Plant}
variant with continuous observation $t_o$.  
We define the observation probability over $t_o$ as uniform on
an interval 10 units wide, centered at $t_s'$.
{\footnotesize
\begin{align}
P(t_o|t_s',\open) = 
\begin{cases}
 (t_o>t_s'-5) \wedge (t_o<t_s'+5) &: 0.1 \\
 \neg((t_o>t_s'-5) \wedge (t_o<t_s'+5)) &: 0 \\
\end{cases}
\end{align}
}
While the restrictions may seem too tight in our problems due to the continuous nature of the state and observation space, there has been no exact solution to any of the two set of problems above. 
\section{DC-POMDP solution}

In a DC-POMDP, the agent does not directly observe the states and thus
must maintain a belief state $b(xd_s) = P(xd_s)$.\footnote{For
compactness we define $xd_s := (x_s,d_s)$ and $xd_o:=(x_o,d_o)$.}  For a
given belief state $\vec{b} = b(\cdot)$, a POMDP policy $\pi$ can be
represented by a tree corresponding to a conditional plan $\beta$.  An
h-step conditional plan $\beta^h$ can be defined recursively in terms
of $(h-1)$-step conditional plans as shown in
Fig.~\ref{fig:cond_plan}.

Our goal is to find a policy $\pi$ that maximizes the value function,
defined as the sum of expected discounted rewards over horizon $h$
starting from initial belief state $\vec{b}$:
\begin{equation}
V^h_\pi(\vec{b}) = E_{\pi} \left[ \sum\nolimits_{t=0}^{h} \gamma^t \cdot r_t \Big| \vec{b}_0 = \vec{b} \right]
\end{equation}
where $r_t$ is the reward obtained at time $t$ and $\vec{b}_0$ is the
belief state at $t=0$.  For finite $h$ and belief state $\vec{b}$, the
optimal policy $\pi$ takes the form of an $h$-step conditional plan
$\beta^h$.  For $h = \infty$, the optimal discounted ($\gamma < 1$)
value function can be approximated arbitrarily closely by using
a sufficiently large, but finite $h$~\cite{kaebling}.

Even when the state is continuous (but the actions and observations
are discrete), 
the optimal POMDP value function for finite horizon $h$ is a piecewise linear
and convex function of the belief state $\vec{b}$~\cite{Perseus_cont} hence 
$V^h$ is given by a maximum over a finite set of
``$\alpha$-vectors'' $\alpha^h_i$:
\begin{equation}
V^h(\vec{b}) = \max\nolimits_{\alpha^h_i \in \Gamma^h} \alpha^h_i \cdot \vec{b}
\end{equation}
Later on when we tackle continuous state \emph{and} observations,
we note that we will dynamically derive an optimal, 
finite partitioning of the observation
space for a given belief state and hence reduce the continuous
observation problem back to a discrete observation problem at every
horizon.

>>>>

@@

\renewcommand{\vec}[1]{\mathbf{#1}} % Bold vectors
\def\argmax{\operatornamewithlimits{arg\max}}

@@

\author
{
Zahra Zamani\\ %\& Scott Sanner\\
ANU \& NICTA\\
Canberra, Australia\\
{\small \texttt{zahra.zamani@anu.edu.au}}\\
%{\small \texttt{first.last@anu.edu.au}}\\
\And
Scott Sanner\\
NICTA \& ANU\\
Canberra, Australia\\
{\small \texttt{scott.sanner@nicta.com.au}}\\
\AND
Pascal Poupart\\
U. of Waterloo\\
Waterloo, Canada\\
{\small \texttt{ppoupart@uwaterloo.ca}}\\
\And 
Kristian Kersting\\
Fraunhofer IAIS \& U. of Bonn\\
Bonn, Germany\\
%{\small \texttt{first.last@ubonn.de???}}\\
%{\small \texttt{first.last@iais.fraunhofer.de}}\\
{\small \texttt{kristian.kersting@iais.fraunhofer.de}}\\
}

@@

\newcommand{\xds}{\mathbf{dx}_s}
\newcommand{\xdsp}{\mathbf{dx}_s'}
\newcommand{\xdo}{\mathbf{dx}_o}
\newcommand{\open}{\mathit{open}}
\newcommand{\close}{\mathit{close}}
\newcommand{\high}{\mathit{high}}
\newcommand{\low}{\mathit{low}}
\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}
\renewcommand{\l}{\langle}
\renewcommand{\r}{\rangle}

\nipsfinalcopy % Uncomment for camera-ready version

@@

\begin{abstract}
%Partially-observable Markov decision processes (POMDPs) provide a
%powerful model for real-world sequential decision-making problems.  
Point-based value iteration (PBVI) methods have
proven extremely effective for finding
(approximately) optimal dynamic programming solutions to
partially-observable Markov decision processes (POMDPs) when a 
set of initial belief states is known.  However, no PBVI work has
provided \emph{exact point-based backups for both continuous state and
observation spaces}, which we tackle in this paper.  Our key insight is
that while there may be an infinite number of observations,
there are only a finite number of continuous observation partitionings
that are relevant for optimal decision-making when a finite, fixed set
of reachable belief states is considered.  To this end, we make two
important contributions: (1) we show how previous exact symbolic
dynamic programming solutions for continuous state MDPs can be
generalized to \emph{continuous state POMDPs with discrete observations}, and
(2) we show how recently developed symbolic integration methods
allow this solution to be extended 
to PBVI for \emph{continuous state and observation POMDPs} with
potentially correlated, multivariate continuous observation spaces.
%We demonstrate a proof-of-concept implementation on power plant regulation.
\end{abstract}

@@

Partially-observable Markov decision processes (POMDPs) are a powerful
modeling formalism for real-world sequential decision-making
problems~\cite{kaebling}.  In recent years, point-based value
iteration methods (PBVI)~\cite{pbvi_jair06,hsvi2,Perseus,gapmin} have proved
extremely successful at scaling (approximately) optimal POMDP
solutions to large state spaces when a set of initial belief states is
known.

While PBVI has been extended to both continuous state and continuous
observation spaces, no prior work has tackled both jointly without sampling.
\cite{Perseus_cont} provides exact point-based backups for continuous
state and discrete observation problems (with approximate sample-based
extensions to continuous actions and observations),
while~\cite{pascal_ijcai05} provides exact point-based backups (PBBs)
for discrete state and continuous observation problems (where
multivariate observations must be conditionally independent).  While
restricted to discrete states, \cite{pascal_ijcai05} provides an
important insight that we exploit in this work: \emph{only a finite
  number of partitionings of the observation space are required to
  distinguish between the optimal conditional policy over a finite set
  of belief states}.

We propose two major contributions:  First, we extend
symbolic dynamic programming for continuous state
MDPs~\cite{sanner_uai11} to POMDPs with discrete 
observations, 
%This provides an expressive and concrete
%instantiation of the framework in~\cite{Perseus_cont} (which only
%abstractly required that integrals were computable) in that it ensures
%the closed-form of all $\alpha$-functions for \emph{all horizons} is a
%symbolic piecewise case statement, even for POMDPs 
\emph{arbitrary} continuous reward and transitions with discrete noise
(i.e., a finite mixture of deterministic transitions).  Second, we
extend this symbolic dynamic programming algorithm to PBVI and the case of
continuous observations (while restricting 
transition dynamics to be piecewise linear with discrete noise, rewards to be
piecewise constant, and observation probabilities and 
beliefs to be uniform) by building 
on~\cite{pascal_ijcai05} to \emph{derive} relevant observation
partitions for potentially correlated, multivariate continuous
observations. % spaces.

%by exploiting the piecewise polynomial integration
%operation of~\cite{sanner_aaai12} and the multivariate symbolic
%maximization technique of~\cite{sanner_uai11}.  We conclude by
%demonstrating our algorithm on a power plant control problem requiring
%bivariate continuous state and observations.

@@

\label{sec:model}

%% TODO: further restrictions on observations!  also discrete problems!
%We assume familiarity with MDPs and
%We introduce discrete and continuous
%partially observable MDPs (DC-POMDPs) as an extension to 
%DC-MDPs~\cite{sanner_uai11}.  
A discrete and continuous
partially observable MDP (DC-POMDP) is a tuple $\langle
\mathcal{S},\mathcal{A},\mathcal{O},\mathcal{T},\mathcal{R},\mathcal{Z},\gamma,h
\rangle$.  States $\mathcal{S}$ are given by vector 
$\xds = (\vec{d}_s,\vec{x}_s) = (
d_{s_1},\ldots,d_{s_n},x_{s_1},\ldots,x_{s_m} )$ where each $d_{s_i}
\in \{ 0,1 \}$ ($1 \leq i \leq n$) is boolean and each
$x_{s_j} \in \mathbb{R}$ ($1 \leq j \leq m$) is continuous.
We assume a finite, discrete action space $\mathcal{A} = \{ a_1,
\ldots, a_r \}$. Observations
$\mathcal{O}$ are given by the vector $\xdo = (\vec{d}_o,\vec{x}_o) = (
d_{o_1},\ldots,d_{o_p},x_{o_1},\ldots,x_{o_q} )$ where each $d_{o_i}
\in \{ 0,1 \}$ ($1 \leq i \leq p$) is boolean and each $x_{o_j} \in
\mathbb{R}$ ($1 \leq j \leq q$) is continuous.

Three functions are required for modeling DC-POMDPs: (1) $\mathcal{T}: \mathcal{S} \times \mathcal{A} \times \mathcal{S} \rightarrow  [ 0, 1 ]$ a Markovian transition model defined as the probability of the next state %$(\vec{d}_s',\vec{x}_s')$ 
given the action and previous state%
%$(\vec{d}_s,\vec{x}_s)$ 
%$p(\vec{d}',\vec{x}'|\cdots,a)$
; (2)  $\mathcal{R}:\mathcal{S}\times\mathcal{A} \rightarrow \mathbb{R}$ a reward function which returns the immediate reward of taking an action in some state; and (3) an observation function defined as $\mathcal{Z} : \mathcal{S} \times \mathcal{A} \times \mathcal{O} \rightarrow [ 0, 1 ]$  which gives the probability of an observation given the outcome of a state after executing an action.  A discount factor $\gamma, \; 0 \leq \gamma \leq 1$ is used to discount rewards $t$ time steps into the future by $\gamma^t$.

We use a dynamic Bayes net (DBN)\footnote{We disallow general 
  synchronic arcs for simplicity of exposition but
  note their inclusion only places restrictions on the variable
  elimination ordering used during the dynamic programming backup
  operation.} to compactly represent the transition model $\mathcal{T}$ over the
factored state variables and we use a two-layer Bayes net to
represent the observation model $\mathcal{Z}$: {\footnotesize
\begin{align}
\mathcal{T}: \;\; &
%p(\vec{d}_s',\vec{x}_s'|\vec{d}_s,\vec{x}_s,a) = 
p(\xdsp|\xds,a) = 
\prod_{i=1}^n p(d_{s_i}'|\xds,a) \prod_{j=1}^m p(x_{s_j}'|\xds, \vec{d}_s',a). \label{eq:trans_model} \\
\mathcal{Z}: \;\; & 
%p(\vec{d}_o,\vec{x}_o|\vec{d}_s,\vec{x}_s,a) = 
p(\xdo|\xdsp,a) = 
\prod_{i=1}^p p(d_{o_i}|\xdsp,a) \prod_{j=1}^q p(x_{o_j}|\xdsp,a). \label{eq:obs_model}
\end{align}}
%% TRANSITION LINEAR, BELIEF, REWARD, OBSERVATION PWC???
Probabilities over \emph{discrete} variables $p(d_{s_i}'|\xds,a)$ and
$p(d_{o_i}|\xdsp,a)$ may condition on both discrete variables and
(nonlinear) inequalities of continuous variables; this is further
restricted to linear inequalities in the case of continuous
observations.  Transitions over \emph{continuous}
variables $p(x_{s_j}'|\xds, \vec{d}_s',a)$ must be deterministic (but
arbitrary nonlinear) piecewise functions;
%encoded using the Dirac $\delta$ function; 
in the case of continuous observations they are further restricted to
be piecewise linear; this permits discrete noise in the continuous 
transitions since they may condition on stochastically sampled
discrete next-state variables $\vec{d}_s'$.
%(hence allowing discrete noise, but not, e.g., Gaussian noise).
Observation probabilities over continuous variables $p(x_{o_j}|\xdsp,a)$ 
only occur in the case of continuous observation and are required to be
piecewise constant (a mixture of uniform distributions); the same
restriction holds for belief state representations.
The reward $R(\vec{d},\vec{x},a)$ may be 
an arbitrary (nonlinear) piecewise function in the case of
deterministic observations and a piecewise constant function in the
case of continuous observations.  
We now provide concrete examples.
%To make this concrete, we now provide examples of both discrete and
%continuous observation DC-POMDPs.

\textbf{Example} \textsc{\bf (Power Plant)~\cite{steam2}} \emph{The steam
generation system of a power plant evaporates feed-water under restricted 
pressure and temperature conditions to turn a stream turbine.
A reward is obtained when electricity is generated from the turbine 
and the steam pressure and temperature are within safe ranges.
Mixing water and steam makes the
respective pressure and temperature observations $p_o \in \mathbb{R}$
and $t_o \in \mathbb{R}$ on the underlying state $p_s \in \mathbb{R}$
and $t_s \in \mathbb{R}$ highly uncertain.  Actions $A = \{ \open, \close \}$
control temperature and pressure by means of a pressure valve.}

We initially present two DC-POMDP variants labeled \textsc{\bf 1D-Power
  Plant} using a single temperature state variable $t_s$.  
The transition and reward are common to both ---
temperature increments (decrements) with a closed (opened) valve, a
large negative reward is given for a closed valve with $t_s$ exceeding
critical threshold $15$, and positive reward is given for a safe, 
electricity-producing state:
% For the transition probability, we use the Dirac function to model the deterministic equations. We can also define stochasticity in the transition model using boolean random variables that are sampled stochastically.  The reward function can be any linear function of the state or action. Going above a threshold temperature (e.g. $T=10$) will cause an explosion in the plant when trying to close the valve which results in the negative reward of $-1000$. Staying below this temperature is safe and will produce electricity an gain the reward of $100$. The reward of an open valve is $-1$.
{\footnotesize
\vspace{-1mm}
\begin{align}
\label{eq:trans}
p(t_s'|t_s,a)= \delta\left[ t_s' - 
\begin{cases}
 (a=\open) &: t_s - 5 \\ 
(a = \close) &: t_s + 7 \\
\end{cases}
\right]
\hspace{2mm}
R(t_s,a) = 
\begin{cases}
 (a=\open) &: -1 \\
(a = \close)\wedge (t_s>15) &: -1000 \\
(a = \close)\wedge \neg(t_s>15) &: 100 \\
\end{cases} 
\end{align}
\vspace{-4mm}
}

%contribution? 
Next we introduce the \textsc{\bf Discrete Obs. 1D-Power Plant} variant where
we define an \emph{observation space with a single discrete binary 
variable} $o \in \mathcal{O} = \{\high,\low\}$:
{\footnotesize
\vspace{-1mm} 
\begin{align}
\hspace{-2.5mm} p(o=\high|t_s',a=\open) = 
\begin{cases}
  t_s' \leq 15 &: 0.9 \\
  t_s' > 15    &: 0.1 \\
\end{cases}
\;\;\;\;
p(o=\high|t_s',a=\close) = 
\begin{cases}
 t_s' \leq 15 &: 0.7 \\
 t_s' > 15    &: 0.3 \\
\end{cases} \label{eq:ex_disc_obs}
\end{align}
\vspace{-4mm}
}
%\\
%p(t_{o_1}|t_s',\close) = 
%\begin{cases}
% \delta\left[ t_s \leq 15 \right] &: 0.1 \\
% \delta\left[ \neg (t_s \leq 15)\right] &: 0.8 \\
%\end{cases}\nonumber
%,
%p(t_{o_2}|t_s',\close) = 
%\begin{cases}
% \delta\left[ t_s \leq 15 \right] &: 0.9 \\
% \delta\left[ \neg (t_s \leq 15)\right] &: 0.2 \\
%\end{cases} \nonumber
%\end{align}

Finally we introduce the \textsc{\bf Cont. Obs. 1D-Power Plant}
variant where we define an \emph{observation space with
a single continuous variable} $t_o$ uniformly distributed on
an interval of 10 units centered at $t_s'$.
{\footnotesize
\vspace{-1mm}
\begin{align}
p(t_o|t_s',a=\open) = U(t_o;t_s' - 5, t_s' + 5) = 
\begin{cases}
 (t_o>t_s'-5) \wedge (t_o<t_s'+5)         &: 0.1 \\
 (t_o \leq t_s'-5) \vee (t_o \geq t_s'+5) &: 0 \\
\end{cases} \label{eq:ex_cont_obs}
\end{align}
\vspace{-4mm}
}
While simple, we note no prior method could perform exact point-based backups for either problem.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 1 - policy tree
\begin{figure}[t!]
%\vspace{-1mm}
\begin{center}
\includegraphics[width=0.4\textwidth]{pics/cond_plan2.pdf}
\hspace{10mm}
\includegraphics[width=0.45\textwidth]{pics/dag2.pdf}
\end{center}
\vspace{-2mm}
\caption{\footnotesize (left) Example conditional plan $\beta^h$ for
discrete observations; (right) example $\alpha$-function for $\beta^h$
over state $b \in \{0,1\}, x \in \mathbb{R}$ in decision
diagram form: the \emph{true} (1) branch is solid, the \emph{false} (0) branch is
dashed.}
\label{fig:cond_plan}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{figure}[t]
%\begin{center}
%\includegraphics[width=0.4\textwidth]{pics/cond_plan2.pdf}
%\end{center}
%\vspace{-3mm}
%\caption{%\footnotesize 
%The optimal value function for \textsc{\bf Power Plant}
%as a decision diagram: 
%the \emph{true} branch is solid, the \emph{false}
%branch is dashed.} 
%\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Value Iteration for DC-POMDPs}

\label{sec:vi}

In a DC-POMDP, the agent does not directly observe the states and thus
must maintain a belief state $b(\xds) = p(\xds)$.  For a
given belief state $\vec{b} = b(\xds)$, a POMDP policy $\pi$ can be
represented by a tree corresponding to a conditional plan $\beta$.  An
h-step conditional plan $\beta^h$ can be defined recursively in terms
of $(h-1)$-step conditional plans as shown in
Fig.~\ref{fig:cond_plan} (left).
Our goal is to find a policy $\pi$ that maximizes the value function,
defined as the sum of expected discounted rewards over horizon $h$
starting from initial belief state $\vec{b}$:
{\footnotesize
\vspace{-1mm}
\begin{equation}
V^h_\pi(\vec{b}) = E_{\pi} \left[ \sum\nolimits_{t=0}^{h} \gamma^t \cdot r_t \Big| \vec{b}_0 = \vec{b} \right]
\end{equation}
\vspace{-4mm}
}

where $r_t$ is the reward obtained at time $t$ and $\vec{b}_0$ is the
belief state at $t=0$.  For finite $h$ and belief state $\vec{b}$, the
optimal policy $\pi$ is given by an $h$-step conditional plan
$\beta^h$.  For $h = \infty$, the optimal discounted ($\gamma < 1$)
value can be approximated arbitrarily closely by 
a sufficiently large $h$~\cite{kaebling}.

Even when the state is continuous (but the actions and observations
are discrete), 
the optimal POMDP value function for finite horizon $h$ is a piecewise linear and
convex function of the belief state $\vec{b}$~\cite{Perseus_cont}, hence 
$V^h$ is given by a maximum over a finite set of
``$\alpha$-functions'' $\alpha^h_i$:
{\footnotesize 
\vspace{-1mm}
\begin{equation}
V^h(\vec{b}) = \max_{\alpha^h_i \in \Gamma^h} \l \alpha^h_i, \vec{b} \r = \max_{\alpha^h_i \in \Gamma^h} \int_{\vec{x}_s} \sum_{\vec{d}_s} \alpha^h_i(\xds) \cdot \vec{b}(\xds) \; d\vec{x}_s 
\end{equation}
\vspace{-4mm}
}

Later on when we tackle continuous state \emph{and} observations,
we note that we will dynamically derive an optimal, 
finite partitioning of the observation
space for a given belief state and hence reduce the continuous
observation problem back to a discrete observation problem at every
horizon.
